---
title: "Simulation and race?"
output:
  html_notebook: default
---

```{r message=FALSE, warning=FALSE}
# Code for plotting densities

getDensity <- function(vector, weights = NULL) {
  
  if (!is.vector(vector)) stop("Not a vector!")
  
  if (is.null(weights)) {
    avg <- mean(vector)
  } else {
    avg <- weighted.mean(vector, weights)
  }
  
  close <- min(which(avg < density(vector)$x))
  x <- density(vector)$x[close]
  if (is.null(weights)) {
    y = density(vector)$y[close]
  } else {
    y = density(vector, weights = weights)$y[close]
  }
  return(data.table(x = x, y = y))
  
}

plotDens <- function(dtx, var1, var2, xPrefix, title, textL = NULL, textT = NULL, weighted = FALSE) {
  
  dt <- copy(dtx)
  
  if (weighted) {
    dt[, nIPW := IPW/sum(IPW)]
    dMarginal <- getDensity(dt[, get(var1)], weights = dt$nIPW)
  } else {
    dMarginal <- getDensity(dt[, get(var1)])
  }
  
  d0 <- getDensity(dt[get(var2)==0, get(var1)])
  d1 <- getDensity(dt[get(var2)==1, get(var1)])

  dline <- rbind(d0, dMarginal, d1)
  
  brk <- round(dline$x, 1)
  
  p <- ggplot(aes(x=get(var1)), data=dt) +
    geom_density(data=dt[get(var2)==0], fill = "#ce682f", alpha = .4) +
    geom_density(data=dt[get(var2)==1], fill = "#96ce2f", alpha = .4)
  
  if (weighted) {
    p <- p + geom_density(aes(weight = nIPW),
                              fill = "#2f46ce", alpha = .8)
  } else p <- p + geom_density(fill = "#2f46ce", alpha = .8)
  
  p <- p +  geom_segment(data = dline, aes(x = x, xend = x, 
                                   y = 0, yend = y), 
                 size = .5, color =  "white", lty=3) +
            annotate(geom="text", x = 75, y = .24, 
               label = title, size = 4, fontface = 2) +
            scale_x_continuous(limits = c(70, 105), 
                       breaks = brk,
                       name = paste(xPrefix, var1)) +
            theme(panel.grid = element_blank(),
                  axis.text.x = element_text(size = 7),
                  axis.title.x = element_text(size = 11)
    )

    if (!is.null(textL))  {
      p <- p + 
        annotate(geom = "text", x = textL[1], y = textL[2], 
                 label = textT[1], size = 3, fontface = 2) +
        annotate(geom = "text", x = textL[3], y = textL[4], 
                 label=textT[2], size = 3, fontface = 2) +
        annotate(geom = "text", x = textL[5], y = textL[6], 
                 label="Population distribution", size = 3, fontface = 2)
    } 
    
    return(p)
}
```

```{r}
library(simstudy)
library(magrittr)
library(broom)
library(ggplot2)
library(gridExtra)
```

```{r}
def <- defData(varname = "x1", formula = "1;12", 
               dist = "uniformInt")
def <- defData(def, varname="y1", formula = "-0.2 + 0.2 * x1", 
               dist="poisson", link="log")

def

```

```{r}
# generate data

set.seed(50)

dt <- genData(1001, def)

head(dt)
```

```{r}
lmfit <- lm(y1~x1,data=dt)
summary(lmfit)
```

```{r, fig.width = 3}
dt[, predlm := predict(lmfit)]

ggplot(dt,aes(y=y1,x=x1)) +
  geom_jitter(width = 0.25, color = "grey60", size = 0.5) +
  geom_line(aes(x=x1, y=predlm), color = "#0932c0", size = 1) +
  scale_x_continuous(breaks = c(1:12)) +
  theme(panel.grid.minor = element_blank())
```

```{r}
glmfit <- glm(y1~x1, data=dt, family=poisson(log))
summary(glmfit)
```

```{r, fig.width = 3}
dt[, predglm := predict(glmfit, type = "response")]
ggplot(dt,aes(y=y1,x=x1)) +
  geom_jitter(width = 0.25, color = "grey60", size = .5) +
  geom_line(aes(x=x1, y=predlm), color = "#0932c0", size = 1, alpha = .3) +
  scale_x_continuous(breaks = c(1:12)) +
  theme(panel.grid.minor = element_blank()) + 
  geom_line(aes(x=x1, y=predglm), color = "#c03b09", size = 1)
```

<!-- Go to slides -->

```{r}
# confounder

defC <- defData(varname = "confounder", formula = .4, dist = "binary")
defC <- defData(defC, varname = "exposure", formula = "10 + 5 * confounder", 
               variance = 4, dist = "normal")
defC <- defData(defC, varname = "outcome", formula = "2*exposure + 5*confounder", 
               variance = 3, dist = "normal")

set.seed(10291)
DTc <- genData(1000, defC)

ggplot(data = DTc, aes(x = exposure, y = outcome)) +
  geom_point(aes(color = factor(confounder)), size = .5) +
  scale_color_manual(values = c("#2d80f0", "#f09d2d"), name = "confounder") +
  theme(panel.grid = element_blank(),
        legend.position = c(.8,.2)) +
  xlim(0,25) +
  ylim(0,60)
```
```{r}
defM <- defData(varname = "exposure", formula = 12, variance = 10, dist = "normal")
defM <- defData(defM, varname = "mediator", formula = "-16 + 1.25 * exposure", 
               dist = "binary", link = "logit")
defM <- defData(defM, varname = "outcome", formula = "2*exposure + 5 * mediator", 
               variance = 3, dist = "normal")

set.seed(10291)
DTm <- genData(1000, defM)

ggplot(data = DTm, aes(x = exposure, y = outcome)) +
  geom_point(aes(color = factor(mediator)), size = .5) +
  scale_color_manual(values = c("#2d80f0", "#f09d2d"), name = "mediator") +
  theme(panel.grid = element_blank(),
        legend.position = c(.8,.2)) +
  xlim(0,25) +
  ylim(0,60)
```
```{r}
tidy(lm(outcome ~ exposure + confounder, data = DTc))[, 1:4]
```

```{r}
tidy(lm(outcome ~ exposure + mediator, data = DTm))[, 1:4]
```

<!-- Go to slides -->

```{r}
defC <- defData(varname = "e", formula = 0, variance = 2, 
                dist = "normal")
defC <- defData(defC, varname = "black", formula = 0.3, 
                dist = "binary")
defC <- defData(defC, varname = "hyp_0", formula = "90 + 5 * black + e", 
                dist = "nonrandom")
defC <- defData(defC, varname = "hyp_1", formula = "80 + 5 * black + e", 
                dist = "nonrandom")
defC <- defData(defC, varname = "ACE", formula = "0.7 - 0.4 * black", 
                dist = "binary")
defC <- defData(defC, varname = "hyp", 
                formula = "hyp_0 * (ACE==0) + hyp_1 * (ACE==1)", 
                dist = "nonrandom")

defC
```

```{r}
set.seed(2017)

dtC <- genData(n = 2000000, defC)
dtC[1:10]
```

```{r, message=FALSE, warning=FALSE, fig.width = 3.25}
p0FULL <- plotDens(dtC, "hyp_0", "black", "Potential outcome", "Full\npopulation",
                   c(86, .24, 98, .22, 81, .06), c("not black", "black"))
p1FULL <- plotDens(dtC, "hyp_1", "black", "Potential outcome", "Full\npopulation")

grid.arrange(p0FULL, p1FULL)
```

#### Marginal distribution of potential outcome changes as distribution of race changes

```{r}
defC <- updateDef(defC, changevar = "black", newformula = 0.50)
defC
```

```{r, warning=FALSE, message=FALSE, fig.width = 3.25}
set.seed(2017)
dtC <- genData(n = 2000000, defC)

p0FULL <- plotDens(dtC, "hyp_0", "black", "Potential outcome", "Full\npopulation")
p1FULL <- plotDens(dtC, "hyp_1", "black","Potential outcome", "Full\npopulation")

grid.arrange(p0FULL, p1FULL)
```

#### Compare potential outcomes with observed outcomes

```{r message=FALSE, warning=FALSE, fig.width = 3.5}
pUnexp <- plotDens(dtC[ACE==0], "hyp", "black","Observed", "No ACE\nonly")
grid.arrange(p0FULL, pUnexp)
```

#### Biases estimating treatment effects

```{r message=FALSE, warning=FALSE, fig.width = 6}
grid.arrange(plotDens(dtC, "hyp_0", "black","Potential outcome", "Population", 
                   c(85, .24, 98, .22, 81, .06), c("not black", "black")),
             plotDens(dtC[ACE==0], "hyp", "black","Observed", "No ACE"),
             plotDens(dtC, "hyp_1", "black","Potential outcome", "Population"),
             plotDens(dtC[ACE==1], "hyp", "black","Observed", "ACE"),
             nrow = 2
)
```

#### Distribution of race in total population

```{r}
dtC[, .(propLis1 = mean(black))]
```

#### Distribution of race in each treatment group

```{r}
dtC[, .(propLis1 = mean(black)), keyby = ACE]
```

#### When treatment is independent of the confounder

```{r}
defC <- updateDef(defC, "ACE", newformula = 0.4) # change data generation
dtC <- genData(n = 2000000, defC)

dtC[, .(propLis1 = mean(black))]            # population/marginal props
```

```{r}
dtC[, .(propLis1 = mean(black)), keyby = ACE] # subgroup proportions
```

```{r, message=FALSE, warning=FALSE, fig.width = 6}
grid.arrange(plotDens(dtC, "hyp_0", "black","Potential outcome", "Population", 
                   c(85, .24, 98, .22, 81, .06), c("not black", "black")),
             plotDens(dtC[ACE==0], "hyp", "black","Observed", "No ACE"),
             plotDens(dtC, "hyp_1", "black","Potential outcome", "Population"),
             plotDens(dtC[ACE==1], "hyp", "black","Observed", "ACE"),
             nrow = 2
)
```

<!-- Go to slides -->

### Experimental

```{r}

# generate correlated data

raceSES <- 0.9
raceStress <- 0.8
sesStress <- 0.8

corMat <- matrix(c(1, raceSES, raceStress, 
                   raceSES, 1, sesStress, 
                   raceStress, sesStress, 1), nrow = 3)

def <- defData(varname = "black", formula = 0.3, dist = "binary")
def <- defData(def, varname = "sesLow", formula = 0.3, dist = "binary")
def <- defData(def, varname = "stress", formula = "-0.5;0.5", dist = "uniform")

set.seed(223984)
dT <- genCorFlex(300, def, corMatrix = corMat) # vary tau 0.1, .5, .8

dT[, cbind(black, sesLow, stress)] %>% cor()
```


```{r}
defA <- defCondition(condition = "sesLow == 0", 
                     formula = "81 + 1*stress",
                     variance = 3, dist = "normal")

defA <- defCondition(defA, condition = "sesLow == 1", 
                     formula = "83 + 10*stress",
                     variance = 3, dist = "normal")

dt <- addCondition(defA, dT, newvar = "hyp")
dt
```


```{r}
pall <- ggplot(data = dt, aes(x = stress, y = hyp)) +
  geom_point()+
  ylim(70, 95) +
  theme(panel.grid.minor = element_blank())

pall
```

```{r}
pall <- ggplot(data = dt, aes(x = stress, y = hyp)) +
  geom_point(aes(color = factor(black))) +
  scale_color_manual(values = c("#bfba10", "#6210bf"), name = "black") +
  ylim(70, 95) +
  theme(panel.grid.minor = element_blank())

pall
```

```{r}
prace <- ggplot(data = dt, aes(x = stress, y = hyp)) +
  geom_point(aes(color = factor(sesLow))) +
  scale_color_manual(values = c("#106dbf", "#bf106d"), name = "sesLow") +
  ylim(70, 95) +
  facet_grid(. ~ factor(black, labels=c("not black", "black"))) +
  theme(panel.grid.minor = element_blank())

prace
```

### Disparities research

```{r}
tidy(lm(hyp ~ black, data = dt))
```

### Further exploration

```{r}
tidy(lmfit <- lm(hyp ~ black + stress, data = dt))
```

```{r}
dt[, pred := predict(lmfit)]

pall + 
  geom_line(aes(x=stress, y=pred, group = black, color = factor(black)), size = 1)

```

```{r}
lmfit <- lm(hyp ~ black * stress, data = dt)
dt[, pred := predict(lmfit)]

pall + 
  geom_line(aes(x=stress, y=pred, group = black, color = factor(black)), size = 1)
```

```{r}
tidy(lmfit)
```

```{r}
prace + 
  geom_line(aes(x=stress, y=pred, group = black), color = "black", size = 1)
```

```{r}
lmfit <- lm(hyp ~ stress, data = dt[black == 1])
dt[black == 1 , pred2 := predict(lmfit)]

prace + 
  geom_line(aes(x=stress, y=pred, group = black), color = "black", size = 1) +
  geom_line(aes(x=stress, y=pred2, group = black), color = "darkgreen", size = 1) 

```

```{r}
tidy(lm(hyp ~ black + stress * sesLow, data = dt))
```


### No correlation

```{r}

# generate correlated data

raceSES <- 0.05
raceStress <- 0.05
sesStress <- 0.05

corMat <- matrix(c(1, raceSES, raceStress, 
                   raceSES, 1, sesStress, 
                   raceStress, sesStress, 1), nrow = 3)

def <- defData(varname = "black", formula = 0.3, dist = "binary")
def <- defData(def, varname = "sesLow", formula = 0.3, dist = "binary")
def <- defData(def, varname = "stress", formula = "-0.5;0.5", dist = "uniform")

set.seed(223984)
dT <- genCorFlex(300, def, corMatrix = corMat) # vary tau 0.1, .5, .8

dT[, cbind(black, sesLow, stress)] %>% cor()
```


```{r}
defA <- defCondition(condition = "sesLow == 0", 
                     formula = "81 + 1*stress",
                     variance = 3, dist = "normal")

defA <- defCondition(defA, condition = "sesLow == 1", 
                     formula = "83 + 10*stress",
                     variance = 3, dist = "normal")

dt <- addCondition(defA, dT, newvar = "hyp")
dt
```


```{r}
pall <- ggplot(data = dt, aes(x = stress, y = hyp)) +
  geom_point()+
  ylim(70, 95) +
  theme(panel.grid.minor = element_blank())

pall
```

```{r}
pall <- ggplot(data = dt, aes(x = stress, y = hyp)) +
  geom_point(aes(color = factor(black))) +
  scale_color_manual(values = c("#bfba10", "#6210bf"), name = "black") +
  ylim(70, 95) +
  theme(panel.grid.minor = element_blank())

pall
```

```{r}
prace <- ggplot(data = dt, aes(x = stress, y = hyp)) +
  geom_point(aes(color = factor(sesLow))) +
  scale_color_manual(values = c("#106dbf", "#bf106d"), name = "sesLow") +
  ylim(70, 95) +
  facet_grid(. ~ factor(black, labels=c("not black", "black"))) +
  theme(panel.grid.minor = element_blank())

prace
```

### Disparities research

```{r}
tidy(lm(hyp ~ black, data = dt))
```

### Further exploration

```{r}
tidy(lmfit <- lm(hyp ~ black + stress, data = dt))
```

```{r}
dt[, pred := predict(lmfit)]

pall + 
  geom_line(aes(x=stress, y=pred, group = black, color = factor(black)), size = 1)

```

```{r}
lmfit <- lm(hyp ~ black * stress, data = dt)
dt[, pred := predict(lmfit)]

pall + 
  geom_line(aes(x=stress, y=pred, group = black, color = factor(black)), size = 1)
```

```{r}
tidy(lmfit)
```

```{r}
prace + 
  geom_line(aes(x=stress, y=pred, group = black), color = "black", size = 1)
```

```{r}
lmfit <- lm(hyp ~ stress, data = dt[black == 1])
dt[black == 1 , pred2 := predict(lmfit)]

prace + 
  geom_line(aes(x=stress, y=pred, group = black), color = "black", size = 1) +
  geom_line(aes(x=stress, y=pred2, group = black), color = "darkgreen", size = 1) 

```

```{r}
tidy(lm(hyp ~ black + stress * sesLow, data = dt))
```

